---
title: "Time Series Project: IPI Food Industries"
output:
  pdf_document:
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_pkgs <- c(
  "readr", "dplyr", "ggplot2", "lubridate",
  "forecast", "tseries", "ellipse", "knitr"
)
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  options(repos = c(CRAN = "https://cloud.r-project.org"))
  install.packages(missing_pkgs)
}

library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(forecast)
library(tseries)
library(ellipse)
library(knitr)
```

# Part I: The data

## 1. What does the chosen series represent?

The dataset is an Industrial Production Index (IPI) for the food industries sector in France, with monthly frequency, seasonally and working-day adjusted (CVS-CJO), base 100 in 2021. The variable is an index and is unit-free. Since the series is already CVS-CJO, we do not apply additional seasonal adjustment.

**Modelling context.**

- As an index (base 100 in 2021), values above (below) 100 indicate production above (below) its 2021 average level for the sector.
- The CVS-CJO correction removes most deterministic seasonal patterns and calendar effects; consequently, we do not expect pronounced remaining seasonality and we focus on non-seasonal ARIMA/ARMA dynamics.
- Because the index is strictly positive over the sample, the analysis can be carried out on $Z_t = \log(Y_t)$; first differences of $Z_t$ have a growth-rate interpretation and often provide variance stabilization.

```{r read-metadata, include=FALSE}
meta_candidates <- list.files("data", pattern = "caract", full.names = TRUE)
if (length(meta_candidates) == 0) {
  stop("Metadata file not found in data/.")
}
meta_path <- meta_candidates[1]
meta <- read_delim(
  meta_path,
  delim = ";",
  col_names = TRUE,
  show_col_types = FALSE
)
names_norm <- iconv(names(meta), from = "", to = "ASCII//TRANSLIT")
pick_col <- function(pattern) {
  idx <- which(grepl(pattern, names_norm, ignore.case = TRUE))
  if (length(idx) == 0) return(NA_character_)
  names(meta)[idx[1]]
}

cols <- c(
  pick_col("idbank"),
  pick_col("mise"),
  pick_col("period"),
  pick_col("unit"),
  pick_col("activ"),
  pick_col("correct"),
  pick_col("indicat"),
  pick_col("zone"),
  pick_col("base")
)
cols <- cols[!is.na(cols)]

meta %>%
  slice(1) %>%
  select(all_of(cols)) %>%
  kable()
```


## 2. Transform the series to make it stationary if necessary

We read the monthly values file, which contains multiple vintages. Each vintage is stored as a pair of columns: a value column followed by a code column (A, P, R, etc.). The row labeled "Mises a jour" gives the update date for each vintage. We select the most recent vintage by default.

```{r read-series, include=FALSE}
values_path <- file.path("data", "valeurs_mensuelles.csv")
raw <- read_delim(
  values_path,
  delim = ";",
  col_names = FALSE,
  show_col_types = FALSE
)

row_label <- raw[[1]]
row_label_norm <- iconv(row_label, from = "", to = "ASCII//TRANSLIT")
row_label_norm <- trimws(row_label_norm)
idx_updates <- which(row_label_norm == "Mises a jour")[1]
idx_period <- which(row_label_norm == "Periode")[1]
if (is.na(idx_updates) || is.na(idx_period)) {
  stop("Could not locate 'Mises a jour' or 'Periode' rows in the values file.")
}

updates_row <- raw[idx_updates, , drop = FALSE]
update_strings <- as.character(unlist(updates_row[-1]))
update_dates <- suppressWarnings(lubridate::dmy_hm(update_strings, quiet = TRUE))

value_cols <- which(!is.na(update_dates)) + 1
vintage_tbl <- tibble(
  vintage_id = seq_along(value_cols),
  update_date = update_dates[!is.na(update_dates)],
  value_col = value_cols
) %>% arrange(update_date)

vintage_tbl %>% tail(6) %>% kable()

latest_vintage <- vintage_tbl %>% slice_max(update_date, n = 1)
chosen_col <- latest_vintage$value_col

series_raw <- raw[(idx_period + 1):nrow(raw), , drop = FALSE] %>%
  transmute(
    period = .data[[names(raw)[1]]],
    value = .data[[names(raw)[chosen_col]]]
  ) %>%
  mutate(
    date = as.Date(paste0(period, "-01")),
    value = readr::parse_number(value, locale = locale(decimal_mark = "."))
  ) %>%
  filter(!is.na(date))

start_year <- year(min(series_raw$date))
start_month <- month(min(series_raw$date))

Y <- ts(series_raw$value, start = c(start_year, start_month), frequency = 12)
Y <- stats::na.omit(Y)

summary(series_raw$value)
```

We consider a log transform if all values are strictly positive. Then we determine the number of differences needed using unit-root tests and verify stationarity with ADF and KPSS.

```{r stationarity, include=FALSE}
use_log <- all(Y > 0, na.rm = TRUE)
Z <- if (use_log) log(Y) else Y

# Suggested differencing from ADF and KPSS
nd_adf <- forecast::ndiffs(Z, alpha = 0.05, test = "adf")
nd_kpss <- forecast::ndiffs(Z, alpha = 0.05, test = "kpss")
d <- max(nd_adf, nd_kpss)

X <- diff(Z, differences = d)

adf_Z <- tseries::adf.test(Z)
kpss_Z <- tseries::kpss.test(Z, null = "Level")

adf_X <- tseries::adf.test(X)
kpss_X <- tseries::kpss.test(X, null = "Level")

list(
  use_log = use_log,
  d_adf = nd_adf,
  d_kpss = nd_kpss,
  d_chosen = d,
  adf_Z_p = adf_Z$p.value,
  kpss_Z_p = kpss_Z$p.value,
  adf_X_p = adf_X$p.value,
  kpss_X_p = kpss_X$p.value
)
```

**Justification.**

- **Log transform.** Since the series is strictly positive over the sample, we work with $Z_t = \log(Y_t)$. Then $\Delta Z_t = \log(Y_t) - \log(Y_{t-1}) \approx (Y_t - Y_{t-1})/Y_{t-1}$.
- **Sample size and frequency.** The series is monthly with `r length(Y)` observations, from `r paste(start(Y), collapse = "-")` to `r paste(end(Y), collapse = "-")`. This satisfies the “at least 100 observations” requirement and is long enough for ARMA identification.
- **Stationarity of $Z_t$.** The ADF test on $Z_t$ has p-value `r signif(adf_Z$p.value, 3)`, so it does not reject a unit root at the 5% level. The KPSS test (level-stationarity null) on $Z_t$ has p-value `r signif(kpss_Z$p.value, 3)`, so it rejects stationarity. Taken together, the tests support the presence of a stochastic trend (non-stationarity) in $Z_t$.
- **Choosing the differencing order $d$.** `forecast::ndiffs()` suggests `r nd_adf` difference(s) under ADF and `r nd_kpss` under KPSS, hence we take the conservative choice $d = \max(d_{ADF}, d_{KPSS}) = `r d`$.
- **Stationarity of $X_t = \nabla^d Z_t$.** After differencing, ADF on $X_t$ yields p-value `r signif(adf_X$p.value, 3)` (reject unit root) while KPSS yields p-value `r signif(kpss_X$p.value, 3)` (do not reject stationarity). This is consistent with $X_t$ being stationary.

We therefore define the stationary modelling series as $X_t = \nabla^{`r d`} \log(Y_t)$.

## 3. Graphical representation before and after transformation

```{r plots-part1, fig.width=7, fig.height=5}
op <- par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
plot(Y, main = "Raw series Y_t", xlab = "Time", ylab = "Index")
plot(X, main = "Stationary series X_t", xlab = "Time", ylab = "Transformed index")
par(op)
```

**Graphical assessment.**

- On the raw index $Y_t$, the level is persistent (slowly moving), which is typical of a non-stationary macro index; large macro events can create visible breaks in level and volatility.
- On $X_t = \Delta \log(Y_t)$, the series fluctuates around a stable mean and shocks are short-lived, which is consistent with a stationary specification.

# Part II: ARMA models

## 4. Pick and estimate an ARMA(p,q) model for $X_t$

We use ACF/PACF diagnostics and an information-criterion search on a small grid to select a parsimonious ARMA model.

**Model identification and selection.**

- **Identification logic.** The ACF/PACF of $X_t$ are used as qualitative guidance: short-range dependence suggests an ARMA with relatively small orders rather than a high-order AR (pure PACF cutoff) or pure MA (pure ACF cutoff).
- **Quantitative selection.** We then compare ARMA$(p,q)$ candidates on a grid ($0\le p,q\le 5$) and select the feasible model minimizing AIC. AIC is preferred here because the goal is forecasting/fit (it penalizes complexity less harshly than BIC).

```{r acf-pacf, fig.width=7, fig.height=5}
op <- par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
forecast::Acf(X, lag.max = 36, main = "ACF of X_t")
forecast::Pacf(X, lag.max = 36, main = "PACF of X_t")
par(op)
```

```{r arma-grid}
max_p <- 5
max_q <- 5

grid <- expand.grid(p = 0:max_p, q = 0:max_q) %>%
  rowwise() %>%
  mutate(
    fit = list(tryCatch(
      Arima(X, order = c(p, 0, q), include.mean = TRUE),
      error = function(e) NULL
    )),
    aic = ifelse(is.null(fit), NA, AIC(fit)),
    bic = ifelse(is.null(fit), NA, BIC(fit))
  ) %>%
  ungroup() %>%
  arrange(aic)

grid %>%
  filter(!is.na(aic)) %>%
  select(p, q, aic, bic) %>%
  head(5) %>%
  kable()

best_row <- grid %>% filter(!is.na(aic)) %>% slice(1)
best_fit <- best_row$fit[[1]]
best_row %>% select(p, q, aic, bic) %>% kable()
```

We select the model with the lowest AIC among feasible candidates, which yields ARMA($p=`r best_row$p`, q=`r best_row$q`$) for $X_t$. The estimated parameters are:

```{r arma-summary}
coef <- best_fit$coef
se <- sqrt(diag(best_fit$var.coef))
coef_tbl <- tibble::tibble(
  term = names(coef),
  estimate = unname(coef),
  std_error = unname(se),
  t_value = unname(coef / se)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

coef_tbl %>% kable()
```

### Diagnostic checks

We validate the fitted model via residual autocorrelation, Ljung-Box test, and a Gaussianity check.

```{r residual-checks, fig.width=7, fig.height=4}
checkresiduals(best_fit, test = FALSE)
```

```{r jb-test, include=FALSE}
res <- residuals(best_fit)
lb_test <- Box.test(res, lag = 24, type = "Ljung-Box", fitdf = sum(best_fit$arma[1:4]))
jb_test <- tseries::jarque.bera.test(res)
```

**Residual diagnostics.**

- **Whiteness.** The Ljung–Box test checks for remaining autocorrelation in residuals. A non-significant p-value supports the ARMA adequacy (no leftover linear dependence).
- **Gaussianity.** The Jarque–Bera test is a strict check for normality. For many macro series, residuals can be non-Gaussian because of outliers (e.g., crisis months), which affects distributional statements (prediction intervals/regions) more than point forecasts.
- **Here (numbers).** Ljung–Box at lag 24 (df adjusted) gives p-value `r signif(lb_test$p.value, 3)`, and Jarque–Bera gives p-value `r signif(jb_test$p.value, 3)`.

Overall, the model is considered acceptable for linear dependence if residual autocorrelation is not significant; the Gaussian assumption used later for the confidence region should be interpreted as an approximation if normality is rejected.

## 5. Write the ARIMA(p,d,q) model for the chosen series

Let $Z_t$ be the transformed series (log if used) and $X_t = \nabla^d Z_t$ be the stationary series. If the selected ARMA model is ARMA$(p,q)$ for $X_t$, then the original series follows an ARIMA$(p,d,q)$ model for $Z_t$. If $Z_t = \log Y_t$, this corresponds to an ARIMA$(p,d,q)$ for the log series, which implies a multiplicative structure on $Y_t$.

**ARIMA representation.** With $d = `r d`$ and the selected ARMA($p=`r best_row$p`, q=`r best_row$q`$) for $X_t = \nabla^d Z_t$, the model can be written as

$$
\phi(B)\,\nabla^{d} Z_t = c + \theta(B)\,\varepsilon_t,\qquad \varepsilon_t \sim \text{i.i.d. } (0,\sigma^2),
$$

where $\phi(B)=1-\phi_1B-\cdots-\phi_pB^p$ and $\theta(B)=1+\theta_1B+\cdots+\theta_qB^q$.
When $Z_t=\log(Y_t)$ and $d=1$, a non-zero mean in $X_t$ corresponds to a **drift** in $\log(Y_t)$, i.e., a constant average growth component.

# Part III: Prediction

Denote $T$ the length of the series. We assume the residuals are Gaussian.

## 6. Confidence region for $(X_{T+1}, X_{T+2})$

Let $\hat{\mu} = (\hat{X}_{T+1}, \hat{X}_{T+2})^\top$ be the 2-step-ahead forecast mean vector and $\hat{\Sigma}$ the 2x2 forecast error covariance matrix. Under Gaussian residuals, the joint forecast error is bivariate normal, so the level-$\alpha$ confidence region is the ellipse

$$
\mathcal{C}_\alpha = \left\{ x \in \mathbb{R}^2 : (x - \hat{\mu})^\top \hat{\Sigma}^{-1} (x - \hat{\mu}) \le \chi^2_{2,\alpha} \right\}.
$$

For an ARMA model, let $\{\psi_j\}_{j\ge 0}$ be the coefficients of the MA($\infty$) representation. Then

- $\operatorname{Var}(X_{T+1} - \hat{X}_{T+1}) = \sigma^2$,
- $\operatorname{Var}(X_{T+2} - \hat{X}_{T+2}) = \sigma^2 (1 + \psi_1^2)$,
- $\operatorname{Cov}(X_{T+1} - \hat{X}_{T+1}, X_{T+2} - \hat{X}_{T+2}) = \sigma^2 \psi_1$,

where $\sigma^2$ is the innovation variance.

**Interpretation.** This region is the multivariate analogue of a univariate prediction interval: it is a joint set such that $\mathbb{P}((X_{T+1},X_{T+2})\in\mathcal{C}_\alpha)\approx \alpha$ under the model. The chi-square threshold comes from the quadratic form of a bivariate normal vector. In practice, if residuals are heavier-tailed than Gaussian, the ellipse may be too small (undercoverage).

## 7. Hypotheses used to obtain this region

1. The ARMA model for $X_t$ is correctly specified.
2. Innovations are i.i.d. Gaussian with variance $\sigma^2$.
3. Model parameters are treated as known (estimation uncertainty ignored for the region).
4. The process is stationary and invertible (so the MA($\infty$) representation exists).

**Role of the assumptions.**

- If the ARMA is misspecified, $\hat\Sigma$ is wrong and the region is not calibrated.
- Gaussianity is what makes the contour exactly elliptical with a chi-square cutoff; without it, ellipses are at best an approximation.
- Treating parameters as known ignores estimation uncertainty and typically makes the region slightly too optimistic; this effect is non-negligible in small samples but tends to shrink as $T$ grows.

## 8. Graphical representation for $\alpha = 95\%$

```{r region-plot, fig.width=6.5, fig.height=5}
alpha <- 0.95

pred <- predict(best_fit, n.ahead = 2)
mu <- as.numeric(pred$pred)

# Extract AR and MA coefficients for psi1
ar <- if (!is.null(best_fit$model$phi)) best_fit$model$phi else numeric(0)
ma <- if (!is.null(best_fit$model$theta)) best_fit$model$theta else numeric(0)
psi1 <- if (length(ar) == 0 && length(ma) == 0) 0 else ARMAtoMA(ar, ma, lag.max = 1)[1]

sigma2 <- best_fit$sigma2
Sigma <- matrix(
  c(sigma2, sigma2 * psi1, sigma2 * psi1, sigma2 * (1 + psi1^2)),
  nrow = 2
)

ell <- as.data.frame(ellipse::ellipse(Sigma, centre = mu, level = alpha))
colnames(ell) <- c("X_T1", "X_T2")

corr12 <- Sigma[1, 2] / sqrt(Sigma[1, 1] * Sigma[2, 2])

p <- ggplot(ell, aes(x = X_T1, y = X_T2)) +
  geom_path(color = "steelblue", linewidth = 1) +
  geom_point(aes(x = mu[1], y = mu[2]), color = "firebrick", size = 2) +
  labs(
    title = "95% joint prediction region for (X_{T+1}, X_{T+2})",
    x = "X_{T+1}",
    y = "X_{T+2}"
  ) +
  theme_minimal()

p
```

**Comments on the 95% region.**

- The forecast-error correlation is `r signif(corr12, 3)`, i.e., **negative**: a positive error at horizon 1 tends to be associated with a negative error at horizon 2 under the fitted ARMA dynamics.
- Using $\rho = \psi_1/\sqrt{1+\psi_1^2}$, this corresponds to $\psi_1 \approx \rho/\sqrt{1-\rho^2}$, hence here $\psi_1 \approx `r signif(corr12 / sqrt(1 - corr12^2), 3)`$.
- The ellipse is therefore tilted downward; its size is governed by the innovation variance $\sigma^2 = `r signif(sigma2, 3)`$ (on the $X_t$ scale) and by $|\psi_1|$.

## 9. Open question

Let $Y_t$ be a stationary time series observed from $t = 1$ to $T$, and suppose $Y_{T+1}$ is observed before $X_{T+1}$. Using $Y_{T+1}$ improves the prediction of $X_{T+1}$ if and only if $Y_{T+1}$ contains incremental predictive information about $X_{T+1}$ beyond the sigma-field generated by $\{X_1, \ldots, X_T\}$. Formally, if

$$
\mathbb{E}[X_{T+1} \mid X_1, \ldots, X_T, Y_{T+1}] \ne \mathbb{E}[X_{T+1} \mid X_1, \ldots, X_T],
$$

then including $Y_{T+1}$ improves the forecast in mean-square error.

A practical test is to estimate an ARMAX-type model for $X_t$ with $Y_{t}$ (or a proxy for $Y_{T+1}$ if only available at $T+1$) and test the significance of the coefficient(s) on $Y$. This can be done via a t-test on the added regressor(s), or a likelihood-ratio test comparing the ARMA model for $X_t$ against the ARMAX model.

**Conditions and practical test.**

- A convenient equivalent condition is that $Y_{T+1}$ is not conditionally independent of $X_{T+1}$ given past $X$: informally, $Y$ must have **incremental predictive content** for $X$.
- In linear/Gaussian settings, this amounts to $Y$ being correlated with the one-step-ahead innovation (forecast error) of $X$ when conditioning on $\mathcal{F}_T=\sigma(X_1,\ldots,X_T)$.
- Practical implementation: fit an ARMAX model $X_t = \text{ARMA terms} + \beta Y_t + \varepsilon_t$ (or a lead/nowcast proxy for $Y_{T+1}$) and test $H_0:\beta=0$ (t-test / Wald) or compare log-likelihoods (LR test). A rejection indicates that observing $Y_{T+1}$ can reduce the MSE of forecasting $X_{T+1}$.

# Appendix: Code

```{r appendix-code, echo=FALSE, results='asis'}
code_chunks <- knitr::knit_code$get()
code_text <- unlist(code_chunks, use.names = FALSE)

cat("```r\n")
cat(paste(code_text, collapse = "\n\n"))
cat("\n```\n")
```
